Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-06 paper.tex

\subsection*{Discrete time Wright-Fisher}
The coalescent is an idealised model and makes many simplifying assumptions,
but it is often surprisingly robust to violations of these
assumptions~\citep{wakeley2012gene}. One situation in which the
model does break down is the combination of large sample size
and with long recombining genomes, where the large
number of recombination events in the recent past results in
more than the biologically possible $2^t$ ancestors in
$t$ diploid generations~\citep{nelson2020accounting}.
This pathological behaviour results in
Identity-by-descent,
long-range linkage disequilibrium (LD), and ancestry patterns deviating from
Wright-Fisher expectations, and the bias grows with larger sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.
Precisely this problem occurs when simulating modern human datasets,
and we have implemented a Discrete Time Wright-Fisher (DTWF) model
in \msprime\ to address the issue. The DTWF simulates backwards in
time generation-by-generation so that each gamete has a unique
diploid parent, and multiple recombinations within a generation results in
crossover events between the same two parental haploid copies.
The method is described in more detail by~\cite{nelson2020accounting}.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/dtwf-perf}
\end{center}
\caption{\label{fig-dtwf-perf} Comparison of Discrete Time Wright-Fisher
(DTWF) simulation performance in \msprime\ and \ARGON\ (Intel Xeon E5-2680 CPU).
We simulated ancestry for a sample of 1000 haploids from a population of 10000,
and report the (A) total CPU time and (B) maximum memory usage for varying
sequence lengths, and a per-base recombination rate of $10^{-8}$. Each
point is the average over $5$ replicate simulations.
We show observations for \ARGON, \msprime's DTWF implementation (``DWTF'')
and a hybrid simulation of 100 generations of the DTWF followed by
the standard coalescent with recombination model (``DTWF + Hudson'').
We ran \ARGON\ with a mutation rate of $0$ and with minimum output options,
to ensure we are measuring only ancestry simulation time.}
\end{figure}

Fig.~\ref{fig-dtwf-perf} shows that \msprime\ simulates the DTWF
more quickly and requires substantially less memory than
\ARGON~\citep{palamara2016argon}, a specialised DTWF simulator.
However, the generation-by-generation approach of the DTWF is less
efficient than the coalescent with recombination when the
number of lineages is significantly less than the population size
(the regime where the coalescent is an accurate approximat),
which usually happens in the quite recent
past~\cite{bhaskar2014distortion}.
We therefore support changing the simulation model during a simulation
so that we can run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}.
Any number of different simulation models can be combined, allowing for the
flexible choice of simulation scenarios.
As the discrete time Wright-Fisher model improves accuracy of genealogical
patterns in the recent past, we can simulate the recent history using this
model and then switch to the standard coalescent to more efficiently simulate
the more ancient history.

>>>>

\subsection*{Discrete time Wright-Fisher}

The assumptions of the classical coalescent model are that both sample size and
the simulated sequence length are small. As modern datasets often include
whole-genome sequences for tens or hundreds of thousands of samples, simulating
such data requires breaking these assumptions. While the coalescent remains a
good approximation to the Wright-Fisher model in the distant past, it can
exhibit large biases in the recent history. Identity-by-descent (IBD),
long-range linkage disequilibrium (LD), and ancestry patterns deviate from
Wright-Fisher expectations, and this bias grows considerably when simulating
larger sample
sizes~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting}.

A backwards-in-time Wright-Fisher model is supported in \msprime, which is
shown to properly capture the expected distributions of IBD and long-range LD,
as well as ancestry patterns in admixed
populations~\citep{nelson2020accounting}. In short, the classical coalescent
draws recombination events from exponential waiting times, which are continuous
in time, and each recombination then splits a segment into two distinct
ancestors. For large sequence lengths, this can dramatically overestimate the
number of genealogical ancestors of a given sample. Instead, the Wright-Fisher
model iterates through discrete generations so that each gamete has a unique
diploid parent, and multiple recombinations within a generation results in
crossover events between the same two parental haploid copies.  This method is
described in more detail in~\cite{nelson2020accounting}.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/dtwf-perf}
\end{center}
\caption{\label{fig-dtwf-perf} Comparison of \msprime's DTWF
simulations with \ARGON. To ensure the comparison with \ARGON\ is
as fair as possible we run it with a mutation rate of 0 and with
the minimum output options (so that it writes out only an empty VCF file).}
\end{figure}

We support changing the simulation model at specified times so that we can
easily run hybrid simulations, as proposed by~\cite{bhaskar2014distortion}.
Any number of different simulation models can be combined, allowing for the
flexible choice of simulation scenarios without severely damaging efficiency.
As the discrete time Wright-Fisher model improves accuracy of genealogical
patterns in the recent past, we can simulate the recent history using this
model and then switch to the standard coalescent to more efficiently simulate
the more ancient history. Fig.~\ref{fig-dtwf-perf} compares the performance of
this approach with \ARGON~\citep{palamara2016argon}, a specialised DTWF
simulator.  The Wright-Fisher approach in \msprime\ is more efficient at
simulating chromosome-scale sequences (larger than $\approx10$Mb), while the
hybrid approach provides large efficiency gains.
